---
title: "Dynamic Report on preprocessing MS-based proteomics data using the msqrob2gui App"
author: "msqrob2gui"
date: "Developed by [statOmics](https://statomics.github.io)"
output:
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
    pdf_document:
      toc: true
      number_sections: true
linkcolor: blue
urlcolor: blue
citecolor: blue

---


```{r libraries}
suppressPackageStartupMessages({
  library(QFeatures)
  library(msqrob2)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})
```

# Import parameters

Initialize all input parameters used in the GUI to import the data.

```{r import-params}
{{importCode}}
```

# Data import

Import the raw data file and transform it into a QFeatures object. Note that, when the input file is in a long format, the intensity column is unique but a run column must be specified. When the input file is in a wide format, we do not need to specify a run column but multiple intensity columns are required (quantCols).

```{r import}
rawFile <- {{rawFileName}}

isLong <- !is.null(runCol) && runCol != "None"

pe_raw <- if (endsWith(rawFile, ".parquet")) {
  arrow::read_parquet(rawFile)
} else {
  data.table::fread(rawFile, check.names = TRUE, integer64 = "double")
}

qf <- if (isLong) {
  QFeatures::readQFeatures(
    assayData = pe_raw,
    fnames    = fnames,
    runCol    = runCol,
    quantCol  = quantCol,
    name      = assayName
  )
} else {
  QFeatures::readQFeatures(
    assayData = pe_raw,
    fnames    = fnames,
    quantCols = which(colnames(pe_raw) %in% quantCols),
    name      = assayName
  )
}
```

# Annotation

Import the annotation file and set it as 'colData' of the QFeatures object.

```{r annotation}
annotFile <- {{annotFileName}}
if (!is.null(annotFile)) {
  annot <- data.table::fread(annotFile, check.names = TRUE, integer64 = "double", stringsAsFactors = TRUE)
  annot <- as.data.frame(annot)
  rownames(annot) <- annot[, 1]
  annot <- annot[rownames(colData(qf)), , drop = FALSE]
  annot <- as.data.frame(lapply(annot, function(col) {
    if (is.character(col)) as.factor(col) else col
  }))
  rownames(annot) <- rownames(colData(qf))
  SummarizedExperiment::colData(qf) <- S4Vectors::DataFrame(annot)
}
```

# Preprocessing parameters

Initialize all input parameters used in the GUI to preprocess the data.

```{r preprocessing-params}
{{preprocessing}}
```

# Preprocessing steps

Missing values are typically encoded as zeros.

```{r preprocessing step 1}
# Step 1: zero to NA
if (doZeroToNA) {
  qf <- QFeatures::zeroIsNA(qf, i = names(qf))
}
```

Apply filtering steps to remove potential false positive identifications.

```{r preprocessing step 2}
# Step 2: filters
if (length(filterList) > 0) {
  filter_formula <- as.formula(paste("~", paste(filterList, collapse = " & ")))
  qf <- QFeatures::filterFeatures(qf, filter_formula)
}
```

Join the multiple assays into a single assay. The 'fcol' parameter controls the feature ID variable used to join the assays.

```{r preprocessing step 3}
# Step 3: join assays (if multiple)
if (length(names(qf)) > 1) {
  qf <- QFeatures::joinAssays(qf, i = names(qf), fcol = fCol, name = nameAssay)
  qf <- qf[, , nameAssay]
}
```

Remove features with a high proportion of missing values across samples.

```{r preprocessing step 4}
# Step 4: filter NA
filterNA_i <- if (!is.null(nameAssay) && nameAssay %in% names(qf)) nameAssay else names(qf)[1]
se <- QFeatures::filterNA(qf[, , filterNA_i], i = filterNA_i, pNA = threshold)[[filterNA_i]]
qf <- QFeatures::addAssay(qf, se, nameFilterNAAssay)
```

Perform log2-transformation.

```{r preprocessing step 5}
# Step 5: log transform
if (doLog) {
  qf <- QFeatures::logTransform(qf, base = 2, i = nameFilterNAAssay, name = nameLogAssay)
}
normInput <- if (!doLog) nameFilterNAAssay else nameLogAssay
```

Apply normalisation to adjust for sample-specific loading differences.

```{r preprocessing step 6}
# Step 6: normalisation
if (normMethod != "None") {
  if (normMethod == "Median of Ratios") {
    qf <- QFeatures::sweep(qf, MARGIN = 2,
                           STATS = msqrob2::nfLogMedianOfRatios(qf, normInput),
                           i = normInput, name = nameNormAssay)
  } else {
    qf <- QFeatures::normalize(qf, method = normMethod, i = normInput, name = nameNormAssay)
  }
}
aggrInput <- if (normMethod == "None") normInput else nameNormAssay
```

Aggregate the features at the protein/peptide level.

```{r preprocessing step 7}
# Step 7: aggregation
if (aggrMethod != "None") {
  aggFun <- if (aggrMethod == "maxLFQ") {
    function(X) iq::maxLFQ(X)$estimate
  } else {
    switch(aggrMethod,
      "medianPolish"  = function(X) MsCoreUtils::medianPolish(X, na.rm = TRUE),
      "robustSummary" = function(X) MsCoreUtils::robustSummary(X, na.rm = TRUE),
      "colMeans"      = function(X) base::colMeans(X, na.rm = TRUE),
      "colMedians"    = function(X) matrixStats::colMedians(X, na.rm = TRUE),
      "colSums"       = function(X) base::colSums(X, na.rm = TRUE)
    )
  }
  qf <- QFeatures::aggregateFeatures(qf, i = aggrInput, name = nameAggrAssay,
                                      fcol = aggrCol, fun = aggFun)
  counts <- SummarizedExperiment::assay(qf[[nameAggrAssay]], "aggcounts")
  a      <- SummarizedExperiment::assay(qf[[nameAggrAssay]])
  a[counts < nprecFilter] <- NA
  SummarizedExperiment::assay(qf[[nameAggrAssay]]) <- a
}
```

Remove proteins with a high proportion of missing values after aggregation. Note that the N-peptides rule introduces missing values for proteins supported by fewer than N peptides, so an additional filtering step is required.

```{r preprocessing step 8}
# Step 8: filter NA post-aggregation (only when aggregation was performed)
if (aggrMethod != "None") {
  se <- QFeatures::filterNA(qf[, , nameAggrAssay], i = nameAggrAssay, pNA = threshold2)[[nameAggrAssay]]
  qf <- QFeatures::addAssay(qf, se, nameFilterNA2Assay)
}
```


# Save the preprocessed QFeatures object

If you want to save the `QFeatures` object, uncomment the line with saveRDS in the R-chunk below. 

Note, that this will overwrite the QFeatures object that you downloaded from the app. 

```{r save}
#saveRDS(qf, "qfeaturesFile.rds")
qf
```

# Session Info

```{r session}
sessionInfo()
```
